---
title: "Quality control checks on long form data for dynamic ICU triage simulation project"
output:
  html_notebook:
    theme: journal
    toc: yes
    toc_depth: 2
    toc_float: yes
  pdf_document: default
---

```{r packages, results=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(survival)
library(gtsummary)
library(survminer)
library(knitr)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Introduction

This script accepts a clean long form data file and performs various quality control checks on it prior to generating transition matrices.

## Load data in and rearrange columns

```{r data_in}
data <- read_csv("wusmbjc_rationing_deidentified2022-02-16.csv")  %>%
  mutate(vent = ifelse(is.na(vent), 0, vent)) %>%
  select(encounter, time_icu, sofa_total, age_years, race, comorbidity, vent, status, everything()) %>%
  arrange(encounter, time_icu) 
```

# Initial sample statistics
```{r initial_icu_sample}
initial_icu_sample <- data %>% 
  group_by(encounter) %>%
  mutate(ever_intubated = max(vent),
         dead = ifelse(status == "died", 1, 0),
         dead = max(dead),
         max_time = max(time_icu),
         end_day = floor(max_time/24),
         week_stay = ifelse(end_day > 6, 1, 0)) %>%
  filter(row_number() ==1) %>%
  rename( sofa_initial = sofa_total) %>%
  mutate(
    sofa_group = cut(sofa_initial, breaks=c(-Inf, 2, 5, 8, Inf), 
                   labels= c("SOFA 0-2", "SOFA 3-5", "SOFA 6-8", "SOFA > 8"))) %>%
                ungroup()
```

## Table 1
```{r table_one}
tbl_summary(initial_icu_sample %>% 
              select(Age = age_years, 
                     "Initial SOFA score" = sofa_initial, 
                     "SOFA distribution" = sofa_group,
                     "Mechanical ventilation at admission" = vent, 
                     "Ever intubated" = ever_intubated,
                     #elix_ahrq,
                     #elix_vw,
                     #charlson,
                     comorbidity,
                     "Race/ethnic group" = race,
                     "Death or discharge to hospice" = dead,
                     "ICU length of stay (days)" = end_day,
                     "ICU stay 7 or more days" = week_stay))
```

## Age distribution
```{r age_distribution}
initial_icu_sample %>%
  ggplot(aes(age_years)) + 
  geom_histogram(color = "black", fill = "white") + 
  labs(x = "Age")
```

## Initial SOFA distribution on ICU admission
```{r sofa_distribution}
initial_icu_sample %>% 
  ggplot(aes(x = sofa_initial)) + 
  geom_histogram(breaks = seq(-0.5,max(initial_icu_sample$sofa_initial)+0.5), color = "black", fill = "white") +
  labs(x = "SOFA") 
```
```{r}
summary(initial_icu_sample$sofa_initial)
```

### Comorbidity score distributions

Need the data to run this code
### Elixhauser AHRQ score distribution
```{r elix_ahrq_score, eval = FALSE}
initial_icu_sample %>% 
  ggplot(aes(x =elix_ahrq )) + 
  geom_histogram(color = "black", fill = "white") +
  labs(x = "Elixhauser Score (AHRQ)") 
```

### VW exlixhuaser
```{r elix_vw_score, eval = FALSE}
initial_icu_sample %>% 
  ggplot(aes(x =elix_vw )) + 
  geom_histogram(color = "black", fill = "white") +
  labs(x = "Elixhauser Score (VW)") 
```


### Charlson Comorbidity Index
```{r charlson_index, eval = FALSE}
initial_icu_sample %>% 
  ggplot(aes(x =charlson)) + 
  geom_histogram(color = "black", fill = "white") +
  labs(x = "Charlson Index") 
```

# Survival analysis
```{r km_by_SOFA}
sofa_initial_fit <- survfit(Surv(end_day, dead) ~ sofa_group, data = initial_icu_sample)
```

```{r survival_by_SOFA}
ggsurvplot(sofa_initial_fit, xlim = c(0, 90), 
           break.time.by = 7,
           xlab = "Days since ICU admission",
           legend  = "top",
           legend.title = "Hour 0 SOFA on admission to ICU",
           legend.labs = c("0-2", "3-5", "6-8", "> 8"),
           palette = "RdBu")
```
```{r survival_summary}
tbl_survfit(sofa_initial_fit, times = c(7, 14, 28, 90))
```

## Statistics by racial/ethnic group 
```{r}
tbl_summary(initial_icu_sample %>% 
              select(Age = age_years, 
                     "Initial SOFA score" = sofa_initial, 
                     "SOFA distribution" = sofa_group,
                     "Mechanical ventilation at admission" = vent, 
                     "Ever intubated" = ever_intubated,
                     #elix_ahrq,
                     #elix_vw,
                     #charlson,
                     comorbidity,
                     "Race/ethnic group" = race,
                      "Survival to ICU discharge" = dead,
                     "ICU length of stay (days)" = end_day,
                     "ICU stay 7 or more days" = week_stay),
            by =  "Race/ethnic group")
```


## Survival to discharge
```{r eval = FALSE}
initial_icu_sample %>%
  group_by(sofa_initial) %>%
  count(dead) %>%
  ggplot(aes(x = sofa_initial, fill = dead, y = n)) + 
  geom_bar(stat = "Identity")
```
```{r icu_mortality_by_initial_SOFA_score}
initial_icu_sample %>%
    mutate(sofa_initial = factor(ifelse(sofa_initial <15, sofa_initial, ">=15"), levels = c(seq(0, 15), ">=15")),
         ) %>%
  group_by(sofa_initial) %>%
  summarise(icu_mortality = mean(dead),
            n = n()) %>%
  mutate(se = sqrt((icu_mortality*(1-icu_mortality))/(n-1)),
         mort_min = icu_mortality - 1.96*se, mort_max = icu_mortality + 1.96*se) %>%
  ggplot(aes(x = sofa_initial, y= 100*icu_mortality, ymin = 100*mort_min, ymax = 100*mort_max)) + 
  geom_bar(stat = "Identity") +
  geom_point() + geom_errorbar() +
  labs(x = "Hour 0 SOFA on ICU admission", y = "Survival to ICU discharge (%)") + lims(y = c(0,100))
```





## ICU mortality by age
```{r icu_mortality_by_age}
initial_icu_sample %>%
  filter(age_years > 18)%>%
  mutate(age_group = cut(age_years, breaks = c(18, 40, 50, 60, 70, 80, Inf))) %>%
  group_by(age_group) %>%
  summarise(icu_mortality = mean(dead),
            n = n()) %>%
  mutate(se = sqrt((icu_mortality*(1-icu_mortality))/(n-1)),
         mort_min = icu_mortality - 1.96*se, mort_max = icu_mortality + 1.96*se) %>%
  ggplot(aes(x = age_group, y= 100*icu_mortality, ymin = 100*mort_min, ymax = 100*mort_max)) + 
  geom_bar(stat = "Identity") +
  geom_point() + geom_errorbar() +
  labs(x = "Age", y = "Survival to ICU discharge (%)") +  lims(y = c(0,100))
```

```{r icu_mortality_by_age_year}
initial_icu_sample %>%
  filter(age_years > 18 & age_years < 100)%>%
  #mutate(age_group = cut(age_years, breaks = c(18, 40, 50, 60, 70, 80, Inf))) %>%
  group_by(age_years) %>%
  summarise(icu_mortality = mean(dead),
            n = n()) %>%
  mutate(se = sqrt((icu_mortality*(1-icu_mortality))/(n-1)),
         mort_min = icu_mortality - 1.96*se, mort_max = icu_mortality + 1.96*se) %>%
  ggplot(aes(x = age_years, y= 100*icu_mortality, ymin = 100*mort_min, ymax = 100*mort_max)) + 
  geom_bar(stat = "Identity") +
  geom_point() + geom_errorbar() +
  labs(x = "Age", y = "Survival to ICU discharge (%)") + lims(y = c(0,100))
```

```{r auc_SOFA}
initial_icu_sample$age_years %>% summary()
```


# Geographic diversity of candidates

Plan to map zip code data to ADI values 



# Possible errors

## Missing hour observations

These are missing hour observations, where presumably the patient had no data. We should decide if we want to fill in the missing hours with duplicated observations

```{r wrong_time_encounters}
wrong_time_ecounters <- data %>%
  arrange(encounter, time_icu) %>%
  mutate(time_check = row_number() -1) %>%
  mutate(wrong_time = ifelse(time_check != time_icu, 1, 0)) %>%
  group_by(encounter) %>%
  mutate(num_wrong_times = sum(wrong_time)) %>%
  filter(num_wrong_times > 0) %>%
  pull(encounter) %>%
  unique()
```



```{r missing_hour_example}
data %>%
  filter(encounter == wrong_time_ecounters[[1]])
```

### Duplicated hours

these are scenarios where a patient had 2 or more SOFA scores in a given hour.We should probably take the maximum
```{r}
duplicated_hours_encounters <- data %>% 
  group_by(encounter, time_icu) %>%
  mutate(hour_repeat = row_number()) %>%
  ungroup() %>%
  filter(hour_repeat > 1) %>% 
  pull(encounter)
```

```{r}
data %>%
  filter(encounter == duplicated_hours_encounters[[1]])
```

## SOFA Coding checks


### SOFA respiratory

### Median PF ratio by coded respiratory SOFA score
```{r}
data %>%
  group_by(sofa_resp) %>%
  filter(is.na(pf_ratio) == FALSE) %>%
  summarise(median_pf = median(pf_ratio))
```


### Min and Max PF ratio by coded respiratory SOFA score
```{r}
data %>%
  group_by(sofa_resp) %>%
  filter(is.na(pf_ratio) == FALSE) %>%
  summarise(min_pf = min(pf_ratio),
            max_pf = max(pf_ratio))
```


### Min and Max SF ratio by coded respiratory SOFA score
```{r}
data %>%
  group_by(sofa_resp) %>%
  filter(is.na(pf_ratio) == TRUE) %>%
  summarise(min_sf = min(sf_ratio, na.rm = TRUE),
            max_sf = max(sf_ratio, na.rm = TRUE))
```


## Underage patients
```{r underage_patients}
underage_patients <- initial_icu_sample %>%
  filter(age_years < 18)

write_csv(underage_patients, "underage_patients.csv")
```
There were `r nrow(underage_patients)` patients under the age of 18 with an average age of `r mean(underage_patients$age_years)`


## Very old patients
```{r too_old_patients}
too_old_patients <- initial_icu_sample %>%
  filter(age_years > 105)

write_csv(too_old_patients, "underage_patients.csv")
```
There were `r nrow(too_old_patients)` patients over the age of 105 with an average age of `r mean(too_old_patients$age_years)`


# To do


1. Add COVID + flag on admission
2. Report actual elixhauser VW, elixhauser AHRQ, and charlson values


