---
title: "Inputs for ICU rationing simulation model"
output: html_notebook
---

```{r}
library(tidyverse)
```

# Read in data set


## Cleaning code specific for Wash U dataset

I've included this here (do not run) as it may be useful in creating the status variables
```{r}
# long_data <- read_csv("covid_sim_anon_2021-02-20.csv")
# 
# icu_enc <- long_data %>%
#   arrange(enc, hours_round) %>%
#   select(enc, hours_round, sofa_final, loc_cat, discharge_dispo,
#          age_years, race, wscore_ahrq) %>%
#   group_by(enc) %>%
#   mutate(in_icu = ifelse(loc_cat == "icu", 1,0),
#          ever_in_icu = max(in_icu)) %>%
#   filter(ever_in_icu == 1) %>%
#   mutate(first_icu = cumsum(in_icu)) %>%
#   filter(first_icu >0) %>%
#   mutate(time_icu = hours_round - min(hours_round),
#          out_icu = ifelse(loc_cat != "icu", 1, 0),
#          out_icu_first = cumsum(out_icu)) %>%
#   filter( out_icu_first <=1) %>%
#   mutate(status = case_when(
#     row_number() < n() ~ loc_cat,
#     loc_cat == "dead" ~ "died",
#     loc_cat == "icu" & discharge_dispo %in% c("dead", "hospice") ~ "died",
#     discharge_dispo == "hospice" ~ "died",
#     loc_cat == "wards" ~ "recovered",
#     is.na(discharge_dispo) == FALSE ~ "recovered")
#     ) %>%
#   select(enc, hours_round, time_icu, sofa_final, loc_cat, discharge_dispo, status,
#          age_years, race, wscore_ahrq) %>%
#   ungroup()
# 
# 
# demographics <- icu_enc %>%
#   group_by(enc) %>%
#   filter(row_number() ==1) %>%
#   select(age_years, race, wscore_ahrq, discharge_dispo)
# 
# major_comorbid <- quantile(demographics$wscore_ahrq, probs = 0.75)
# 
# severe_comorbid <-  quantile(demographics$wscore_ahrq, probs = 0.90)
# 
# demographics <- demographics %>%
#   mutate(comorbidity = case_when(
#     wscore_ahrq < major_comorbid ~ "None",
#     wscore_ahrq < severe_comorbid ~ "Major",
#     is.na(wscore_ahrq)== FALSE ~ "Severe"
#   ))
# 
# icu_enc <- icu_enc %>%
#   left_join(demographics)%>%
#   select(enc, time_icu, sofa_final, loc_cat, status, age_years, race, comorbidity) %>%
#   mutate(race = case_when(
#     race %in% c("White", "Black") ~ race,
#     TRUE ~ "Other"
#   ),
#   race = factor(race)
#   )

```


## Details on cleaning to apply to your data

* Use SaO2/FiO2 if PaO2/FiO2 is unavailable 
* Code 0 for missing SOFA score components
* "Carryforward" values of the SOFA components, i.e. if the renal SOFA was 1 on admission it remains 1 until a new lab/urine output result changes it.
* The comorbidity categorization in this dataset uses an AHRQ elixhauser of < `r major_comorbid` for none, `r major_comorbid` - `r severe_comorbid` for major, and > `r severe_comorbid` for severe. This is not set in stone obviously and we will need to standardize othis across datasets
* Deaths that occur after transfer to the wards but prior to discharge are coded as recoveries
* Transfers to hospice from the ICU are coded as deaths, regardless of exactly when/where the patient dies 
* Do not need to filter out patients who are still in the hospital at the end of follow-up. Can use their censored data in the transition matrices
* Please include race **and** ethnicity data, coded as Non-Hispanic Black, Non-Hispanic White, Hispanic, and Other.

## Load your data and update your file name
```{r}
#icu_data <- read_csv("your_data.csv")

#institution_name <- "Wash_U"

icu_enc %>%
  head()
```



# Create inputs

## Age categories and duration of ICU stay
```{r}
ages <- c(18, 45, 60, 75, Inf)
days <- c(0, 3, 7, Inf)
racial_groups <- levels(icu_enc$race)
```


## SOFA transition matrix
```{r}
sofa_trans_matrix <- function(data, min_age, max_age, start_day, stop_day, comorbid_cat){
  M <- data %>%
    group_by(enc) %>%
    filter(age_years >= min_age & age_years < max_age &
             time_icu >= start_day*24 & time_icu <= stop_day*24 &
             comorbidity == comorbid_cat)%>%
    mutate(sofa_final = ifelse(sofa_final >= 17, "[17,INF)", as.character(sofa_final)),
           next_state = case_when(
             row_number() == n()-1 & lead(status) %in% c("recovered", "died") ~lead(status),
             TRUE ~lead(sofa_final)),
           sofa_final = factor(sofa_final, levels = c(as.character(seq(0,17)), "[17,INF)")),
           next_state = factor(next_state, levels = c(as.character(seq(0,17)), "[17,INF)", "recovered", "died"))
      ) %>%
    filter(row_number() != n()) %>%
    ungroup() %>%
    group_by(sofa_final) %>%
    count(next_state, .drop = FALSE) %>%
    spread(key = "next_state", value = "n")
  
  
  write_csv(M, paste0(institution_name,"/sofa_transitions/sofa_transition_age_",min_age,"-", max_age, "_intubation_", start_day, "-", stop_day, "_comorbids_", comorbid_cat, ".csv"))
}

dir.create(institution_name)
dir.create(paste0(institution_name,"/sofa_transitions"))
           
for (a in seq(1, length(ages)-1)) {
  for (d in seq(1, length(days)-1)) {
    for (c in c("None", "Major", "Severe")) {
      sofa_trans_matrix(icu_enc, ages[[a]], ages[[a+1]], days[[d]], days[[d+1]], c)
    }
  }
  
}
```
Note that the SOFA transition matrix does **not** depend on race, just age, comorbidities, and duration of ICU stay

## Initial SOFA score by race, age, and comorbidity

```{r}

sofa_initial_distribution <- function(data, min_age, max_age, comorbid_cat, cur_race){
   M <- data %>% 
     group_by(enc) %>%
    filter(row_number() == 1 &
             age_years >= min_age & age_years < max_age &
             comorbidity == comorbid_cat &
             race == cur_race)%>%
    mutate(sofa_initial = ifelse(sofa_final >= 17, "[17,INF)", as.character(sofa_final)),
           sofa_initial = factor(sofa_initial, 
                                 levels = c(as.character(seq(0,17)),"[17,INF)"))) %>% 
  ungroup() %>%
  count(sofa_initial,
        .drop = FALSE)
   
     write_csv(M, paste0(institution_name,"/sofa_initial_distributions/sofa_initial_age_",min_age,"-", max_age,"_comorbids_", comorbid_cat, "_race_", cur_race, ".csv"))
}

dir.create(paste0(institution_name, "/sofa_initial_distributions"))

for (r in racial_groups) {
  for (a in seq(1, length(ages)-1)) {
    for (c in c("None", "Major", "Severe")) {
      sofa_initial_distribution(icu_enc, ages[[a]], ages[[a+1]], c, r)
    }
  } 
}

```


## Comorbidity distribution for each age group by race

```{r}
dir.create(paste0(institution_name, "/comorbidities_by_age"))
comorbids_age <- function(data, age_list, cur_race){
  M <- data %>% 
  group_by(enc) %>%
    filter(row_number() == 1,
           race == cur_race)%>%
    ungroup() %>%
  mutate(age_cat = cut(age_years, breaks = age_list, right = FALSE),
         comorbidity = factor(comorbidity, 
                              levels = c("None", "Major", "Severe"))) %>%
  group_by(age_cat) %>%
  count(comorbidity, .drop = FALSE) %>%
    spread(key = "comorbidity", value = "n")
  
       write_csv(M, paste0(institution_name,"/comorbidities_by_age/comorbidities_by_age_", cur_race, ".csv"))
}

for (r in racial_groups) {
  comorbids_age(icu_enc, ages, r)
}
```


## Age distribution

```{r}
dir.create(paste0(institution_name, "/age_distribution"))
age_distribution <- function(data, cur_race){

  M <- data %>% 
  group_by(enc) %>%
    filter(row_number() == 1 & race == cur_race)%>%
    ungroup() %>%
  mutate(age_years = cut(age_years, breaks =  c(seq(18, 90), Inf), right = FALSE)) %>%
  count(age_years, .drop = FALSE)
  
       write_csv(M, paste0(institution_name, "/age_distribution/age_distribution_", cur_race, ".csv"))
}
for (r in racial_groups) {
  age_distribution(icu_enc, r)
}

```

## Racial distribution

```{r}
dir.create(paste0(institution_name, "/racial_distribution"))
race_distribution <- function(data){
  M <- data %>% 
  group_by(enc) %>%
    filter(row_number() == 1)%>%
    ungroup() %>%
  count(race, .drop = FALSE)
  
       write_csv(M, paste0(institution_name, "/racial_distribution/racial_distribution_.csv"))
}

race_distribution(icu_enc)
```

Observed racial distribution




